---
title: "Combining multiple data tables"
author: "Math 285, Adam Loy"
date: "Day 08"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
```


## Your turn: Space launches

The **agencies** and **launches** data sets can be loaded using the below code chunk, and you can find the code books in [this Tidy Tuesday GitHub Repository](https://github.com/rfordatascience/tidytuesday/tree/master/data/2019/2019-01-15).

```{r message=FALSE}
agencies <- read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-01-15/agencies.csv")
launches <- read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-01-15/launches.csv")
```


### Task 1

Use an appropriate join statement to merge the agencies and launches data tables such that all rows from the launches data table are retained.
```{r}
left_join(x= launches, y = agencies, by = "state_code")
```




### Task 2

Calculate the number of SpaceX launches by year.
```{r}
left_join(x= launches, y = agencies, by = "state_code")
```



### Task 3

Create a time series plot of the number of SpaceX launches by year.

