---
title: "Introduction to Data Wrangling with dplyr"
author: "Math 285, Adam Loy"
date: "Day 07"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(babynames)
```

This .Rmd file contains the "Your turn" question text. Please complete these questions as we work through these concepts in class.

## babynames

```{r}
babynames
```


## select()

Which of these is NOT a way to select the `name` and `n` columns together?
select(babynames, ends_with("n"))"
```{r}
select(babynames, -c(year, sex, prop))
select(babynames, name:n)
select(babynames, starts_with("n"))
select(babynames, ends_with("n"))
```

## filter()

Use `filter()` with the logical operators to extract:

1. All of the names where `prop` is greater than or equal to 0.08

2. All of the babies named "Sea"

3. All of the names that have a missing value for `n`


```{r}
#1
babynames %>% filter(prop >= 0.08)
#2
babynames %>% filter(name == "Sea")
#3
filter(babynames, is.na(n))

babynames %>% filter(name =="Noam")
```



## filter() with Boolean operators

Use Boolean operators to alter the code below to return only the rows that contain:

- Girls named Sea

- Names that were used by exactly 5 or 6 children in 1880

- Names that are one of Acura, Lexus, or Yugo

```{r}
filter(babynames, name == "Sea" | name == "Anemone")
babynames %>%  filter(name == "Sea", sex = "F") 

babynames %>% filter (n == "5" | n == "6") 
filter(babynames, year == 1880 & (n==5 | n==6))

babynames %>% filter(name == "Acura" | name == "Lexus" | name == "Yugo")
filter(babynames, name %in% c("Acura", "Lexus", "Yugo"))
```



## arrange()

- Arrange babynames by `n`. 

- Add `prop` as a second (tie breaking) variable to arrange by.

- Can you tell what the smallest value of `n` is?

```{r}
arrange(babynames, n)
arrange(babynames, n, prop)
arrange(babynames, desc(n))
```



## Descending order

- Use `desc()` to find the names with the highest `prop`

- Then, use `desc()` to find the names with the highest `n`.

```{r}
arrange(babynames, desc(n))
arrange(babynames, desc(prop))
```



## filter() + select() + arrange()

1. Filter `babynames` to just girls born in 2015

2. Select the `name` and `n` columns from the result

3. Arrange those columns so that the most popular names appear near
the top.

```{r}
babynames %>% filter(sex == "F", year == "2015") %>% select(name, n) %>% arrange(desc(n))
```



## Putting it all together

1. Trim `babynames` to just the rows that contain your name
and your sex

2. Trim the result to just the columns that will appear in
your graph (not strictly necessary, but useful practice)

3. Plot the results as a line graph with `year` on the x axis and `prop` on the y axis

4. Add an informative title and polish the axis labels

```{r}
babynames %>% 
  filter(name == "Noam", sex == "M") %>% 
  ggplot(aes(x = year, y = prop)) + 
  geom_line() + 
  labs(x = "Year",
       y = "Proportion of Kids Named Noam",
       title = "Proportion of Children Born Named Noam Between 1880 and 2017")
```



## summarize()

Use `summarize()` to compute three statistics about the data:

1. The first (minimum) year in the dataset

2. The last (maximum) year in the dataset

3. The total number of children represented in the data

```{r}
summarize(babynames,
          min = min(year), 
          max = max(year), 
          sum = sum(n))


```



## summarize() 2


Extract the rows where `name == "Khaleesi"`. Then use `summarize()` and a summary functions to find:

1. The total number of children named Khaleesi 

2. The first year Khaleesi appeared in the data

```{r}
babynames %>% 
  filter(name == "Khaleesi") %>% 
  summarize(Khaleesis = sum(n))
```



## Grouped summaries

- Use `group_by()`, `summarize()`, and `arrange()` to display the ten most popular names of all time

- Compute popularity as the total number of children of a single sex given a name

```{r}
babynames %>% 
  group_by(name) %>%
  summarize(nName = sum(n)) %>%
  arrange(desc(nName)) %>%
  head(n = 10)
```



## Ranking baby names

Use `min_rank()` and `mutate()` to rank each row in `babynames` from largest `prop` to smallest `prop`.

```{r}

babynames %>% 
  mutate(rank = min_rank(desc(prop))) %>%
  arrange(rank)

#to fully embrace the pipe operator
babynames %>%
  mutate (ranking = prop 
          %>% desc() 
          %>% min_rank()) %>%
  
```


## Grouped rankings

- Compute each name's rank *within its year and sex*.

- Then compute the median rank for *each combination of name and sex*, and arrange the results from highest median rank to lowest.

```{r}
babynames %>% 
  group_by(year, sex) %>%
  mutate(rank = min_rank(desc(prop))) %>%
  arrange(rank, year)


babynames %>% 
  group_by(year, sex) %>%
  mutate(rank = min_rank(desc(prop))) %>%
  group_by(name, sex) %>%
  mutate(median = median(rank)) %>%
  arrange(median)

#embracing pipe operator
babynames %>% 
  group_by(year, sex) %>%
  mutate(rank = prop %>%
           desc() %>%
           min_rank()) %>%
  group_by(name, sex) %>%
  mutate(median = rank %>%
           median()) %>%
  arrange(median)

```

