---
title: "Changing themes and scales"
author: "Math 285, Adam Loy"
date: "Day 05"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview

Today in class we'll explore how to change the theme (i.e. non-data ink) in your ggplot2 graphics. Along the way, we'll also see how to set aesthetic parameters and work with a few scales.


## Example 1: Calories in Food

Last class, we saw a hideous bar chart displaying the number of calories in five foods. Our first task is to create a polished version of this plot.


### (a) Create a bar chart displaying the calories by food, coloring the bars by `bacon`.

The data shown last class are created in the below code chunk.

```{r message=FALSE, warning=FALSE}
library(tidyverse)
calories <- tribble(
  ~food, ~calories, ~bacon,
  "French Fries", 607, FALSE,
  "Potato chips", 542, FALSE,
  "Bacon", 533, TRUE,
  "Pizza", 296, FALSE,
  "Chili Dog", 260, FALSE
)
```

```{r 1a solution}
ggplot(data = calories, aes(x=food, y = calories, fill=bacon)) + geom_col() + labs(x= "Food Type", y = "Calories", title = "Callories by Food", subtitle = "Color Shows Whether Bacon or Not")

ggplot(data = calories, aes(x=food, y = calories, fill=bacon)) + geom_bar(stat = "identity") + labs(x= "Food Type", y = "Calories", title = "Callories by Food", subtitle = "Color Shows Whether Bacon or Not")
```

### (b) Reorder the bars by caloric content.

We can use the `fct_inorder()` command to reorder the bars within our ggplot2 code since the rows are ordered. 

```{r 1b solution}
base <- ggplot(data = calories, aes(x=fct_inorder(food), y = calories, fill=bacon)) + geom_col() + labs(x= "Food Type", y = "Calories", title = "Calories by Food", subtitle = "Color Shows Whether Bacon or Not", caption = "Ordered by Calories")
base
```

Other helpful functions for reordering factors for plotting include `fct_reorder()`, `fct_infreq()`, and `fct_rev()`. These functions are part of the `forcats` package which is loaded by the `tidyverse`.


### (c) Remove the axis labels, since they are redundant.

One approach is to set their `labs()` to `NULL`:

```{r 1c solution}
ggplot(data = calories, aes(x=fct_inorder(food), y = calories, fill=bacon)) + geom_col() + labs(NULL)
```

Another approach is to set the appropriate element of the `theme()` to `element_blank()`:

```{r 1c solution2}
base + 
  theme(axis.title = element_blank())
base <- base + theme(axis.title = element_blank())
```



### (d) Remove the legend, since we are simply calling attention to bacon.

We can edit the appearance of the legend using a variety of arguments within the `theme()` (see `?theme` for the full list). To remove the legend, we set `legend.position = "none"` (Other options for the  position include`"left"`, `"right"`, `"bottom"`, `"top"`, or a two-element numeric vector)

```{r 1d solution}
base + 
  theme(legend.position = "none")
base <- base + theme(legend.position = "none")
```

### (e) Change the scale for fill color.

The default fill colors are not ideal, especially since we are simply drawing attention to bacon. To change the scale, we appeal to a `scale_fill_<method>()` function. A few options include:

- `scale_fill_brewer()`
- `scale_fill_grey()`
- `scale_fill_viridis_c()`
- `scale_fill_manual()`

We'll use the last option today. 

With `scale_fill_manual()`, you can manually specify named colors for a factor in the order of the levels:

```{r e solution1}
base + 
  scale_fill_manual(values = c("gray40", "red2"))
base <- 
# Change the fill colors to "gray40" and "red2"
```

You can also use other formats, like RGB

```{r e solution2}
# Change the fill colors to (R: 163, G: 163, B: 163) and (R: 192, G: 66, B: 68)
base + 
  scale_fill_manual(values = c(
    rgb(163, 163, 163, maxColorValue = 255), 
    rgb(192, 66, 68, maxColorValue = 255)
    ))
```

or hex codes

```{r e solution3}
# Change the fill colors to "#DFDCE3" and "#FC4A1A"
base + 
  scale_fill_manual(values = c("#DFDCE3", "#FC4A1A"))
base <- base + scale_fill_manual(values = c("#DFDCE3", "#FC4A1A"))
```

### (f) Remove the gray background.

The background adds a lot of non-data ink, so many people prefer to remove it. Remember that to remove an element of a theme, we set it to `element_blank()`.

```{r f solution}
# Remove the grid's background color
base + 
  theme(panel.background = element_blank())
base <- base + theme(panel.background = element_blank())
```

### (g) Remove tick marks on the axes.


```{r g solution}
# Remove the axis tick marks
base + 
  theme(axis.ticks = element_blank())
base <- base + theme(axis.ticks = element_blank())
```


## (h) Add guidelines (Tufte style)

Edward Tufte didn't like grids on plots, but he acknowledged that they helped interpret plots. For bar charts, he made a compromise, adding the grid only to the bars. In ggplot2, this is accomplished by adding white horizontal lines via `geom_hline()`.

```{r h solution}
# Add white horizontal lines at the labeled values on the y-axis
base + 
  geom_hline(yintercept = c(200, 400, 600), 
                  color = "white", 
                  lwd = 1)
```

## (i) Direct labels: remove y-axis text and annotate the bars

An alternative is to directly label the bars. We can achieve this by adding a layer via `geom_text()`. Notice that we only needed to specify the `label` here because the other aesthetic mappings are _inherited_ from the base layer.

```{r i solution}
# Add text to directly label the bars
base + 
  geom_text(
    aes(label = calories),
    vjust = -0.3,
    color = "black"
  ) +
  theme(axis.text.y = element_blank()) 
```



## Example 2: Polishing the Economist graphic

On Friday, you tried to recreate a graphic exploring the average net cost of US colleges by income quintile. Below is my version on this graphic based on the tools we had on Friday. 

```{r ex2 template}
scorecard <- read_csv("https://github.com/cmsc205/data/raw/master/ScorecardSmallNarrow.csv")

econ <- ggplot(scorecard, aes(x = ADM_RATE*100, y = net_cost/1000)) +
  geom_point(alpha = 0.2, color='steelblue2') +
  geom_smooth(se = FALSE, color = 'black') +
  facet_grid(fct_rev(CONTROL) ~ income_group) +
  labs(x = "Admissions rate, %", y = "Net cost per year, $`000",
       title = "Scorecard",
       subtitle = "Average net cost per year at U.S. colleges",
       caption = "*Colleges with at least 1,000 undergrads")
econ

```


Use what you have learned about themes to polish this plot to more-closely resemble the plot from *The Economist*. Feel free to explore the `ggthemes` package and the help file for `theme()`.

```{r ex2 solution}
# Keep editing the above plot in an effort to maatch the Economist's theme
ggplot(data = scorecard) +
  geom_point(aes(x = ADM_RATE*100, y = net_cost/1000), alpha = 0.2, color='deepskyblue3') +
  facet_grid(CONTROL~income_group, as.table = FALSE) +
  geom_smooth(aes(x = ADM_RATE*100, y = net_cost/1000), se = FALSE, color = 'black') +
  labs(
    title = "Scorecard",
    subtitle = "Average net cost per year at US colleges, by income quintile",
    x = "Admissions rate, %",
    y = "Net cost per year, $1000",
    caption = "*Colleges with at least 1,000 undergrads"
  )
```
```{r}
econ +
  theme(
    plot.background = element_rect(fill = "gray93"),
    panel.background = element_rect(fill = "white"),
    plot.title = element_text(face = "bold"),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_line(linetype = 3, color = "gray50"),
    axis.line.x = element_line(),
    strip.background = element_rect(fill = "gray75") +
    scale_y_continuous(breaks = seq(0,50,by=10)  ))
```

